In-silico molecular enrichment and clearance of the human intracranial space

### 2025-06-11 paper/main.tex
Address minor points from Reviewer #3

% What is the big picture problem and why is it important? 
The mechanisms underlying molecular transport within the intracranial
space are fundamental to human brain health and
function. Neurodegenerative diseases such as Alzheimer's, Parkinson's,
and Huntington's disease are all associated with abnormal accumulation
of protein aggregates together with alterations in transport and
clearance characteristics~\cite{rasmussen2018glymphatic,
  harrison2020impaired, eide2023plasma, liu2024glymphatic}. Moreover,
sleep and conversely sleep-deprivation play a definite yet enigmatic
role in modulating molecular enrichment and
clearance~\cite{xie2013sleep, eide2021sleep, eide2022altered,
  miao2024brain, hauglund2025norepinephrine}. In the last decade,
established theories have been challenged by new findings revealing a
greater degree of molecular movement and
exchange~\cite{iliff2012paravascular, ringstad2017glymphatic,
  louveau2017understanding, proulx2021cerebrospinal,
  bohr2022glymphatic}, as well as substantial variability in
enrichment and clearance between individuals and between patient
cohorts~\cite{ringstad2018brain, eide2021direction, eide2021impaired,
  eide2022altered}. These observations provide distinct opportunities
(and challenges) for personalized medicine e.g.~for tailored
intrathecal delivery of chemotherapies\cite{lohela2022glymphatic}
such as methodextrate in acute lymphoblastic leukemia
patients\cite{kadan2009comparison}, and for early diagnostics of
impaired brain clearance\cite{eide2021clinical, van2024human}. In
spite of their importance, our understanding of these mechanisms is
incomplete with open challenges and significant debate -- in part
relating to the translation of knowledge between scales, species,
experimental protocols, clinical cohorts, and individuals.

Perivascular pathways along the brain surface and within the brain
parenchyma have long been hypothesized to serve a designated role in
this context~\cite{rennels1985evidence, zhang1990interrelationships,
  ichimura1991distribution, carare2008solutes, iliff2012paravascular,
  foley2012realtime, hannocks2018molecular, van2024caa}. Recently, Eide and
Ringstad~\cite{eide2024functional} and Yamamoto et
al~\cite{yamamoto2024perivascular} demonstrated that perivascular
spaces (PVSs) define preferential pathways for molecular transport in
humans, with delayed periarterial enrichment in dementia
subtypes~\cite{eide2024functional}. Perivascular flow of cerebrospinal
fluid (CSF) clearly contributes to this transport, and is inherently
associated with vascular pulsations~\cite{hadaczek2006perivascular,
  iliff2013cerebral, bedussi2018paravascular, mestre2018flow,
  boster2023artificial, hirschler2024region}. Fluid mechanics
considerations point at intracranial pressure differences and shorter
wavelength vascular wall pulsations as drivers of directional net flow
and convection in the PVS~\cite{bilston2003arterial, rey2018pulsatile,
  daversin2020mechanisms, kedarasetti2020functional,
  gjerde2023directional, nozaleda2024arterial}, while longer waves
such as the pulse wave primarily contribute to oscillatory flow and
dispersion~\cite{asgari2016glymphatic, sharp2019dispersion, thomas2019fluid,
  kedarasetti2020arterial, troyetsky2021dispersion, 
  martinac2021phase}. In the bigger picture, CSF is produced by the
choroid plexus~\cite{damkier2013cerebrospinal,
  steffensen2018cotransporter, liu2020direct}, pulsates through the
ventricular system, cisterns, and subarachnoid space (SAS) in
synchrony with cardiac, respiratory, and neural
waves~\cite{greitz1993pulsatile, wagshul2011pulsating,
  sweetman2011cerebrospinal, fultz2019coupled, vinje2019respiratory,
  eide2021direction, causemann2022human, williams2023neural,
  zimmermann2023total}, and drains via the dural sinuses, meningeal
lymphatics, cranial nerves, or other efflux
pathways~\cite{proulx2021cerebrospinal}. However, how these
physiological factors and physical mechanisms integrate to enhance or
impair human intracranial molecular transport over larger spatial
scales and longer time scales remain unknown. A related key question
is to what extent surface PVSs are separated from the SAS by
structural barriers~\cite{zhang1990interrelationships,
  weller2005microscopic, bedussi2017paravascular,
  pizzo2018intrathecal, mestre2022periarteriolar,
  mollgard2023mesothelium, smets2024perivascular, eide2024functional},
and in turn to what extent such structural compartmentalization is a
prerequisite for effective perivascular transport.

@@

%% Summary of findings
We have presented a high-fidelity in-silico model of molecular
enrichment and clearance in human intracranial spaces, enabling
tailored predictions of the influence of CSF space and (peri)vascular
morphology, physiological factors such as cardiac and respiratory
pulsatility, as well as of pathological conditions such as enlarged
PVSs in the SAS. A key observation is that the balance between CSF
production and intracranial pulsatility significantly shapes the
large-scale features of molecular enrichment, with e.g.~ventricular
tracer reflux after intrathecal injection in the absence of CSF
production, and preference towards anterior brain regions with reduced
pulsatility. Indeed, the simulated distribution patterns cover a
substantial range of the individual variations observed clinically
\cite{ringstad2018brain}. In terms of perivascular pathways, we find
that even moderate CSF flow on the order of 0.1--1.0 \textmu m/s in
the surface PVSs results in earlier tracer enrichment around major
cerebral arteries. This effect is more pronounced -- with up to a
three-hour difference in time of arrival -- with mean PVS flow rates
of approximately 10 \textmu m/s. Our models of CSF flow in
perivascular networks, based on first principles and asymptotic
analysis, predict net flow of such magnitudes and both antegrade and
retrograde flow within the perivascular network, occurring with or
without pressure differences at the network ends. The PVS may thus
function as highways facilitating rapid transport, even in the absence
of structural barriers, though a sharp concentration difference
between the PVS and surrounding tissues is unlikely without such
barriers. Finally, dilated PVS in the SAS will cause a substantial reduction in
cardiac- and vasomotion-driven flow velocities, which in turn impedes
perivascular transport.

>>>>

% What is the big picture problem and why is it important? 
The mechanisms underlying molecular transport within the intracranial
space are fundamental to human brain health and
function. Neurodegenerative diseases such as Alzheimer's, Parkinson's,
and Huntington's disease are all associated with abnormal accumulation
of protein aggregates together with alterations in transport and
clearance characteristics~\cite{rasmussen2018glymphatic,
  harrison2020impaired, eide2023plasma, liu2024glymphatic}. Moreover,
sleep and conversely sleep-deprivation play a definite yet enigmatic
role in modulating molecular enrichment and
clearance~\cite{xie2013sleep, eide2021sleep, eide2022altered,
  miao2024brain, hauglund2025norepinephrine}. In the last decade,
established theories have been challenged by new findings on molecular
movement and exchange~\cite{iliff2012paravascular,
  ringstad2017glymphatic, louveau2017understanding,
  proulx2021cerebrospinal, bohr2022glymphatic}, including substantial
variability between individuals and between patient
cohorts~\cite{ringstad2018brain, eide2021direction, eide2021impaired,
  eide2022altered}. These observations provide distinct opportunities
(and challenges) for personalized medicine e.g.~for tailored
intrathecal delivery of chemotherapies\cite{lohela2022glymphatic} such
as methodextrate in acute lymphoblastic leukemia
patients\cite{kadan2009comparison}, and for early diagnostics of
impaired brain clearance\cite{eide2021clinical, van2024human}. In
spite of their importance, our understanding of these mechanisms is
incomplete with open challenges and significant debate -- in part
relating to the translation of knowledge between scales, species,
experimental protocols, clinical cohorts, and individuals.

Perivascular pathways along the brain surface and within the brain
parenchyma have long been hypothesized to serve a designated role in
this context~\cite{rennels1985evidence, zhang1990interrelationships,
  ichimura1991distribution, carare2008solutes, iliff2012paravascular,
  foley2012realtime, hannocks2018molecular, van2024caa}. Recently, Eide and
Ringstad~\cite{eide2024functional} and Yamamoto et
al~\cite{yamamoto2024perivascular} demonstrated that perivascular
spaces (PVSs) define preferential pathways for molecular transport in
humans, with delayed periarterial enrichment in dementia
subtypes~\cite{eide2024functional}. Perivascular flow of cerebrospinal
fluid (CSF) clearly contributes to this transport, and is inherently
associated with vascular pulsations~\cite{hadaczek2006perivascular,
  iliff2013cerebral, bedussi2018paravascular, mestre2018flow,
  boster2023artificial, hirschler2024region}. Fluid mechanics
considerations point at intracranial pressure differences and shorter
wavelength vascular wall pulsations as drivers of directional net flow
and convection in the PVS~\cite{bilston2003arterial, rey2018pulsatile,
  daversin2020mechanisms, kedarasetti2020functional,
  gjerde2023directional, nozaleda2024arterial}, while longer waves
such as the pulse wave primarily contribute to oscillatory flow and
dispersion~\cite{asgari2016glymphatic, sharp2019dispersion, thomas2019fluid,
  kedarasetti2020arterial, troyetsky2021dispersion, 
  martinac2021phase}. In the bigger picture, CSF is produced by the
choroid plexus~\cite{damkier2013cerebrospinal,
  steffensen2018cotransporter, liu2020direct}, pulsates through the
ventricular system, cisterns, and subarachnoid space (SAS) in
association with cardiac, respiratory, and neural
waves~\cite{greitz1993pulsatile, wagshul2011pulsating,
  sweetman2011cerebrospinal, fultz2019coupled, vinje2019respiratory,
  eide2021direction, causemann2022human, williams2023neural,
  zimmermann2023total}, and drains via the dural sinuses, meningeal
lymphatics, cranial nerves, or other efflux
pathways~\cite{proulx2021cerebrospinal}. However, how these
physiological factors and physical mechanisms integrate to enhance or
impair human intracranial molecular transport over larger spatial
scales and longer time scales remain unknown. A related key question
is to what extent surface PVSs are separated from the SAS by
structural barriers~\cite{zhang1990interrelationships,
  weller2005microscopic, bedussi2017paravascular,
  pizzo2018intrathecal, mestre2022periarteriolar,
  mollgard2023mesothelium, smets2024perivascular, eide2024functional},
and in turn to what extent such structural compartmentalization is a
prerequisite for effective perivascular transport.

@@

%% Summary of findings
We have presented a high-fidelity in-silico model of molecular
enrichment and clearance in human intracranial spaces, enabling
tailored predictions of the influence of CSF space and (peri)vascular
morphology, physiological factors such as cardiac and respiratory
pulsatility, as well as of pathological conditions such as enlarged
PVSs in the SAS. A key observation is that the balance between CSF
production and intracranial pulsatility significantly shapes the
large-scale features of molecular enrichment, with e.g.~ventricular
tracer reflux after intrathecal injection in the absence of CSF
production, and preference towards anterior brain regions with reduced
pulsatility. Indeed, the simulated distribution patterns cover a
substantial range of the individual variations observed clinically
\cite{ringstad2018brain}. In terms of perivascular pathways, we find
that even moderate CSF flow on the order of 0.1--1.0 \textmu m/s in
the surface PVSs results in earlier tracer enrichment around major
cerebral arteries. This effect is more pronounced -- with up to a
three-hour difference in time of arrival -- with mean PVS flow rates
of approximately 10 \textmu m/s. Our models of CSF flow in
perivascular networks, based on first principles and asymptotic
analysis, predict net flow of such magnitudes and both antegrade and
retrograde flow within the perivascular network. The PVS may thus
function as highways facilitating rapid transport, even in the absence
of structural barriers, though a sharp concentration difference
between the PVS and surrounding tissues is unlikely without such
barriers. Finally, dilated PVS in the SAS will cause a substantial reduction in
cardiac- and vasomotion-driven flow velocities, which in turn impedes
perivascular transport.
